---
title: "PBMC Sc_rna seq analysis"
output: html_document
date: "2025-07-15"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,   # show code
  warning = FALSE,  # hide warnings
  message = FALSE   # hide package messages (optional)
)
```


```
```{r}
#install.packages("Seurat")
#install.packages("patchwork")
```

```{r}
setwd("/Users/tharun kota/OneDrive/Desktop/UCD/Trimester_3/Internship")
```

```{r}
library(dplyr)
library(Seurat)
library(patchwork)

# Load the PBMC dataset
pbmc.data <- Read10X(data.dir = "pbmc3k_filtered_gene_bc_matrices/filtered_gene_bc_matrices/hg19/")
# Initialize the Seurat object with the raw (non-normalized data).
pbmc <- CreateSeuratObject(counts = pbmc.data, project = "pbmc3k", min.cells = 3, min.features = 200)
pbmc
```
```{r}
# The [[ operator can add columns to object metadata. This is a great place to stash QC stats
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
```

```{r}
# Visualize QC metrics as a violin plot
VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```
```{r}
# FeatureScatter is typically used to visualize feature-feature relationships, but can be used
# for anything calculated by the object, i.e. columns in object metadata, PC scores etc.

plot1 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```
```{r}
pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
```

```{r}
pbmc <- NormalizeData(pbmc, normalization.method = "LogNormalize", scale.factor = 10000)
```
```{r}
pbmc <- FindVariableFeatures(pbmc, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(pbmc), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(pbmc)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2
```
```{r}
all.genes <- rownames(pbmc)
pbmc <- ScaleData(pbmc, features = all.genes)
```
```{r}
pbmc <- RunPCA(pbmc, features = VariableFeatures(object = pbmc))
```
```{r}
VizDimLoadings(pbmc, dims = 1:2, reduction = "pca")
```
```{r}
DimPlot(pbmc, reduction = "pca") + NoLegend()
```
```{r}
DimHeatmap(pbmc, dims = 1:15, cells = 500, balanced = TRUE)
```
```{r}
ElbowPlot(pbmc)
```
```{r}
pbmc <- FindNeighbors(pbmc, dims = 1:10)
pbmc <- FindClusters(pbmc, resolution = 0.5)
```
```{r}
# Look at cluster IDs of the first 5 cells
head(Idents(pbmc), 5)
```
```{r}
pbmc <- RunUMAP(pbmc, dims = 1:10)
```
```{r}
# note that you can set `label = TRUE` or use the LabelClusters function to help label
# individual clusters
DimPlot(pbmc, reduction = "umap")
```
```{r}
saveRDS(pbmc, file = "pbmc_tutorial.rds")
```

```{r}
# find markers for every cluster compared to all remaining cells, report only the positive
# ones
pbmc.markers <- FindAllMarkers(pbmc, only.pos = TRUE)
pbmc.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)
```



```{r}
VlnPlot(pbmc, features = c("MS4A1", "CD79A"))
```
```{r}
# you can plot raw counts as well
VlnPlot(pbmc, features = c("NKG7", "PF4"), slot = "counts", log = TRUE)
```

```{r}
FeaturePlot(pbmc, features = c("MS4A1", "GNLY", "CD3E", "CD14", "FCER1A", "FCGR3A", "LYZ", "PPBP",
    "CD8A"))
```
```{r}
pbmc.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10
DoHeatmap(pbmc, features = top10$gene) + NoLegend()
```
```{r}
new.cluster.ids <- c("Naive CD4 T", "CD14+ Mono", "Memory CD4 T", "B", "CD8 T", "FCGR3A+ Mono",
    "NK", "DC", "Platelet")
names(new.cluster.ids) <- levels(pbmc)
pbmc <- RenameIdents(pbmc, new.cluster.ids)
DimPlot(pbmc, reduction = "umap", label = TRUE, pt.size = 0.5) + NoLegend()
```


```{r}
pbmc$annotations <- Idents(pbmc)
```

```{r}
devtools::install_github("saeyslab/nichenetr")
```

```{r}
library(nichenetr) # Please update to v2.0.4
library(Seurat)
library(SeuratObject)
library(tidyverse)
```
 
 Nichenet Analysis

```{r}
seuratObj <- pbmc

seuratObj@meta.data %>% head()

seuratObj <- UpdateSeuratObject(seuratObj)
```

```{r}
seuratObj <- alias_to_symbol_seurat(seuratObj, "human")
```

```{r}

seuratObj@meta.data$celltype %>% table() 
DimPlot(seuratObj, reduction = "umap")
```

```{r}
lr_network <- readRDS("C:/Users/tharun kota/OneDrive/Desktop/UCD/Trimester_3/Internship/lr_network_human_21122021.rds")
ligand_target_matrix <- readRDS("C:/Users/tharun kota/OneDrive/Desktop/UCD/Trimester_3/Internship/ligand_tf_matrix_nsga2r_final.rds")
weighted_networks <- readRDS("C:/Users/tharun kota/OneDrive/Desktop/UCD/Trimester_3/Internship/weighted_networks_nsga2r_final.rds")

lr_network <- lr_network %>% distinct(from, to)
head(lr_network)

ligand_target_matrix[1:5,1:5] # target genes in rows, ligands in columns


head(weighted_networks$lr_sig) # interactions and their weights in the ligand-receptor + signaling network

head(weighted_networks$gr) # interactions and their weights in the gene regulatory network

```

```{r}
receiver = "Memory CD4 T"
expressed_genes_receiver <- get_expressed_genes(receiver, seuratObj, pct = 0.05)
```

```{r}
all_receptors <- unique(lr_network$to)  
expressed_receptors <- intersect(all_receptors, expressed_genes_receiver)

potential_ligands <- lr_network %>% filter(to %in% expressed_receptors) %>% pull(from) %>% unique()
```

```{r}
sender_celltypes <- "CD14+ Mono"

# Use lapply to get the expressed genes of every sender cell type separately here
list_expressed_genes_sender <- sender_celltypes %>% unique() %>% lapply(get_expressed_genes, seuratObj, 0.05)
expressed_genes_sender <- list_expressed_genes_sender %>% unlist() %>% unique()

potential_ligands_focused <- intersect(potential_ligands, expressed_genes_sender) 

# Also check 
length(expressed_genes_sender)

length(potential_ligands)

length(potential_ligands_focused)

```

```{r}
DE_table_receiver <-  FindMarkers(object = seuratObj,
                                  ident.1 = receiver, ident.2 =       sender_celltypes,
                                  min.pct = 0.05) %>% rownames_to_column("gene")

geneset_oi <- DE_table_receiver %>% filter(p_val_adj <= 0.05 & abs(avg_log2FC) >= 0.25) %>% pull(gene)
geneset_oi <- geneset_oi %>% .[. %in% rownames(ligand_target_matrix)]
```

```{r}
background_expressed_genes <- expressed_genes_receiver %>% .[. %in% rownames(ligand_target_matrix)]

length(background_expressed_genes)

length(geneset_oi)

```

```{r}
ligand_activities <- predict_ligand_activities(geneset = geneset_oi,
                                               background_expressed_genes = background_expressed_genes,
                                               ligand_target_matrix = ligand_target_matrix,
                                               potential_ligands = potential_ligands)

#ligand_activities <- ligand_activities %>% arrange(-aupr_corrected) %>% mutate(rank = rank(desc(aupr_corrected)))
ligand_activities <- ligand_activities %>%
  arrange(desc(aupr_corrected)) %>%
  mutate(rank = rank(-aupr_corrected))
ligand_activities
```

```{r}
p_hist_lig_activity <- ggplot(ligand_activities, aes(x=aupr_corrected)) + 
  geom_histogram(color="black", fill="darkorange")  + 
  geom_vline(aes(xintercept=min(ligand_activities %>% top_n(30, aupr_corrected) %>% pull(aupr_corrected))),
             color="red", linetype="dashed", size=1) + 
  labs(x="ligand activity (PCC)", y = "# ligands") +
  theme_classic()

p_hist_lig_activity
```


```{r}
best_upstream_ligands <- ligand_activities %>% top_n(30, aupr_corrected) %>% arrange(-aupr_corrected) %>% pull(test_ligand)
```

```{r}
library(dplyr)
```

```{r}
vis_ligand_aupr <- ligand_activities %>% filter(test_ligand %in% best_upstream_ligands) %>%
  column_to_rownames("test_ligand") %>% dplyr::select(aupr_corrected) %>% arrange(aupr_corrected) %>% as.matrix(ncol = 1)

(make_heatmap_ggplot(vis_ligand_aupr,
                     "Prioritized ligands", "Ligand activity", 
                     legend_title = "AUPR", color = "darkorange") + 
    theme(axis.text.x.top = element_blank()))  
```

```{r}
active_ligand_target_links_df <- best_upstream_ligands %>%
  lapply(get_weighted_ligand_target_links,
         geneset = geneset_oi,
         ligand_target_matrix = ligand_target_matrix,
         n = 100) %>%
  bind_rows() %>% drop_na()

nrow(active_ligand_target_links_df)
head(active_ligand_target_links_df)
```

```{r}
#increasing the cutoff will result in a denser or sparser heatmap, respectively.

active_ligand_target_links <- prepare_ligand_target_visualization(
  ligand_target_df = active_ligand_target_links_df,
  ligand_target_matrix = ligand_target_matrix,
  cutoff = 0.10) 

nrow(active_ligand_target_links)
head(active_ligand_target_links)

order_ligands <- intersect(best_upstream_ligands, colnames(active_ligand_target_links)) %>% rev()
order_targets <- active_ligand_target_links_df$target %>% unique() %>% intersect(rownames(active_ligand_target_links))

vis_ligand_target <- t(active_ligand_target_links[order_targets,order_ligands])

make_heatmap_ggplot(vis_ligand_target, "Prioritized ligands", "Predicted target genes",
                    color = "purple", legend_title = "Regulatory potential") +
  scale_fill_gradient2(low = "whitesmoke",  high = "purple")
  

# keep top 30 targets by column sum of regulatory potential
keep_targets <- names(sort(colSums(vis_ligand_target), decreasing = TRUE))[1:30]
vis_small <- vis_ligand_target[, keep_targets, drop = FALSE]

make_heatmap_ggplot(
  vis_small,
  "Prioritized ligands",
  "Top predicted target genes",
  color = "purple",
  legend_title = "Regulatory potential"
) +
  scale_fill_gradient2(low = "whitesmoke", high = "purple") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 7),
    axis.text.y = element_text(size = 8)
  )
```

```{r}
ligand_receptor_links_df <- get_weighted_ligand_receptor_links(
  best_upstream_ligands, expressed_receptors,
  lr_network, weighted_networks$lr_sig) 
```

```{r}
vis_ligand_receptor_network <- prepare_ligand_receptor_visualization(
  ligand_receptor_links_df,
  best_upstream_ligands,
  order_hclust = "both") 

(make_heatmap_ggplot(t(vis_ligand_receptor_network), 
                     y_name = "Ligands", x_name = "Receptors",  
                     color = "mediumvioletred", legend_title = "Prior interaction potential"))
```


